---
title: "Explore and Preprocess - Version 1"
output: html_notebook
---

```{r setup}
library(dplyr)
library(arsenal)
library(ggplot2)
library(caret)
library(InformationValue)
library(pROC)
library(ROCR)
library(mediation)
library(mice)
library(GGally)

data_raw = read.csv("./data__raw/main_raw_Oct_17.csv", stringsAsFactors=T)
```

A couple of decisions:
* Using CXCL10 LOD from the raw table

<!-- Function for transforming Ct values to viral load: -->
<!-- ```{r} -->
<!-- ct_to_vl = function(ct){ -->
<!--   if (is.na(ct)){ -->
<!--     return (0) -->
<!--   }else{ -->
<!--     vl = 2^(-(ct - 45)) -->
<!--     return (vl) -->
<!--   } -->
<!-- } -->
<!-- ``` -->


# View Data

```{r}
data_raw
```
# Filter and Rename columns

Current names:
```{r}
names(data_raw)
```
Filtering and renaming:

```{r}
data_filt = data_raw %>%
  # remove stuff we don't need
  dplyr::select(-c(
    RVP.data, # it contains badly coded data from the other columns
    Ct.value, # it contains badly coded data from the other columns
    Patient.Class, # dont think we need that
    CXCL10..Ella., # We use the LOD one
    IL.1B..pg.mL.# We use the LOD one
    )) %>%
  rename(study_code = Study.code,
       any_virus = Any.virus,
       rvp_adeno_ct = Adeno,
       rvp_rv_ct = RV,
       rvp_hmpv_ct = HMPV,
       rvp_cov229_ct = CoV.229E.OC43,
       rvp_covnl63_ct = CoV.NL63,
       rvp_flu_a_ct = Flu.A,
       rvp_piv_ct = PIV,
       rvp_rsv_ct = RSV,
       any_bacteria = Any.Bacteria,
       bac_pneu_ct = S..pneumoniae.ct,
       bac_influ_ct = H..influenzae.ct,
       bac_catarr_ct = M..catarrhalis.ct,
       albumin_ct = Albumin.Ct,
       gender = Gender,
       age = Age..decimals.,
       # cxcl10_ella = CXCL10..Ella.,
       cxcl10 = CXCL10..pg.mL..with.LOD,
       # il_1b = IL.1B..pg.mL.,
       il_1b = IL.1B..pg.mL..with.LOD,
       sars_interpret = SARS.CoV.2..COVID.19..Interpretation,
       sars_roche_orf_ct = Roche.cobas...ORF1ab,
       sars_roche_e_ct = Roche.cobas...E,
       sars_cepheid_n2_ct = Cepheid.Xpert...N2,
       sars_cepheid_e_ct = Cepheid.Xpert...E,
       sars_cdc_ct = CDC.N1..Foxman.Lab.,
       patient_present = Clinical.Presentation.After.Chart.Review.Emergency.Use..ARI..PARI..NRI.
       )
data_filt
```
# Preprocessing subset

Preprocessing the following domains:
* Demo, status, and outcome biomarkers
* Virus + COVID
* Bacterial

## Demo, status, and outcome biomarkers

1. Log transform outcome biomarkers CXCL10 and IL-1b
2. Coding `symptomatic` based on `patient_present`: 
  * ARI and PARI -> Y
  * PREOP and NRI -> N

```{r}
data_pp_demo = data_filt %>%
  dplyr::select(study_code, gender, age, patient_present, cxcl10, il_1b ) %>%
  mutate(cxcl10_log = log(cxcl10), .after = cxcl10) %>%
  mutate(il_1b_log = log(il_1b), .after = il_1b) %>%
  mutate(symptomatic = ifelse(patient_present == "ARI" | patient_present == "PARI", "Y", "N"), .after = patient_present) %>%
  mutate(symptomatic = as.factor(symptomatic)) %>%
  dplyr::select(-patient_present)

data_pp_demo
```
## Process Covid

```{r}
data_pp_sars = data_filt %>%
  dplyr::select(study_code, sars_interpret, sars_roche_orf_ct, sars_roche_e_ct, sars_cepheid_n2_ct,
                sars_cepheid_e_ct, sars_cdc_ct)
data_pp_sars
```
### Code the test interpretation

Coding `any_sars` based on `sars_interpret`: 
* Negative and Not Detected -> N
* Positive and Presumptive Positive -> Y

```{r}
data_pp_sars = data_pp_sars %>%
  mutate(any_sars = ifelse(sars_interpret == "Negative" | sars_interpret == "Not Detected", "N", "Y"), .after = sars_interpret) %>%
  mutate(any_sars = as.factor(any_sars)) %>%
  dplyr::select(-sars_interpret)

```

### Unify SARS CT

What we want:
* One variable for SARS Ct
* A pair of variables where one has NAs replaces with Inf and a boolean for Ct missing

#### Simple (mean)

Calculate the mean
```{r}
data_pp_sars = data_pp_sars %>%
  rowwise() %>%
  mutate(sars_ct = mean(c(sars_roche_orf_ct, sars_roche_e_ct, sars_cepheid_n2_ct, sars_cepheid_e_ct, sars_cdc_ct), na.rm = T), 
         sars_ct = ifelse(is.na(sars_ct), NA, sars_ct),
         sars_ct_imp = ifelse(is.na(sars_ct), Inf, sars_ct),
         sars_ct_is_na = as.factor(ifelse(is.na(sars_ct), "Y", "N"))) 
data_pp_sars

```
#### Filter

```{r}
data_pp_sars = data_pp_sars %>% 
  dplyr::select(study_code, any_sars, sars_ct, sars_ct_imp, sars_ct_is_na)
```

## Process RVP

Steps:
1. Create a column that contains max/min/mean Ct -> NA Cts is NA in these
2. For each CT, generate an imputed column and a column encoding NA status:
3. Create a new column that just contains a y or n for any respiratory virus present (not SARS)
4. Create a new column that just contains a y or n for any respiratory virus categories
5. create a column that contains number of rvp infections


```{r}
names(data_filt)
```

Create a column that contains max/min/mean Ct -> NA Cts is NA in these
```{r warning=FALSE}
data_pp_vir = data_filt %>% 
  dplyr::select(study_code, any_virus, rvp_adeno_ct, rvp_rv_ct, rvp_hmpv_ct, rvp_cov229_ct, rvp_covnl63_ct, rvp_flu_a_ct, rvp_piv_ct, rvp_rsv_ct) 

data_pp_vir = data_pp_vir %>%
  # average and max/min ct
  rowwise() %>%
  # calculate
  mutate(rvp_avg_ct = mean(c(rvp_adeno_ct, rvp_rv_ct, rvp_hmpv_ct, rvp_cov229_ct, rvp_covnl63_ct, rvp_flu_a_ct, rvp_piv_ct, rvp_rsv_ct), na.rm = T),
         rvp_max_ct = max(c(rvp_adeno_ct, rvp_rv_ct, rvp_hmpv_ct, rvp_cov229_ct, rvp_covnl63_ct, rvp_flu_a_ct, rvp_piv_ct, rvp_rsv_ct), na.rm = T),
         rvp_min_ct = min(c(rvp_adeno_ct, rvp_rv_ct, rvp_hmpv_ct, rvp_cov229_ct, rvp_covnl63_ct, rvp_flu_a_ct, rvp_piv_ct, rvp_rsv_ct), na.rm = T)) %>%
  # replace - inf
  mutate(rvp_avg_ct = ifelse(rvp_avg_ct < 0, NA, rvp_avg_ct),
       rvp_max_ct = ifelse(rvp_max_ct < 0, NA, rvp_max_ct),
       rvp_min_ct = ifelse(rvp_min_ct -- Inf, NA, rvp_min_ct))
data_pp_vir
```

For each CT, generate an imputed column and a column encoding NA status:
```{r}
data_pp_vir = data_pp_vir %>% mutate( 
    rvp_adeno_ct_imp = ifelse(is.na(rvp_adeno_ct), Inf, rvp_adeno_ct),
    rvp_adeno_ct_is_na = ifelse(is.na(rvp_adeno_ct), "Y", "N"),
    
    rvp_rv_ct_imp = ifelse(is.na(rvp_rv_ct), Inf, rvp_rv_ct),
    rvp_rv_ct_is_na = as.factor(ifelse(is.na(rvp_rv_ct), "Y", "N")),
    
    rvp_hmpv_ct_imp = ifelse(is.na(rvp_hmpv_ct), Inf, rvp_hmpv_ct),
    rvp_hmpv_ct_is_na = as.factor(ifelse(is.na(rvp_hmpv_ct), "Y", "N")),
    
    rvp_cov229_ct_imp = ifelse(is.na(rvp_cov229_ct), Inf, rvp_cov229_ct),
    rvp_cov229_ct_is_na = as.factor(ifelse(is.na(rvp_cov229_ct), "Y", "N")),
    
    rvp_covnl63_ct_imp = ifelse(is.na(rvp_covnl63_ct), Inf, rvp_covnl63_ct),
    rvp_covnl63_ct_is_na = as.factor(ifelse(is.na(rvp_covnl63_ct), "Y", "N")),
    
    rvp_flu_a_ct_imp = ifelse(is.na(rvp_flu_a_ct), Inf, rvp_flu_a_ct),
    rvp_flu_a_ct_is_na = as.factor(ifelse(is.na(rvp_flu_a_ct), "Y", "N")),
    
    rvp_piv_ct_imp = ifelse(is.na(rvp_piv_ct), Inf, rvp_piv_ct),
    rvp_piv_ct_is_na = as.factor(ifelse(is.na(rvp_piv_ct), "Y", "N")),
    
    rvp_rsv_ct_imp = ifelse(is.na(rvp_rsv_ct), Inf, rvp_rsv_ct),
    rvp_rsv_ct_is_na = as.factor(ifelse(is.na(rvp_rsv_ct), "Y", "N"))
    ) 
```

Create a new column that just contains a y or n for any respiratory virus present (not SARS)
```{r}
data_pp_vir = data_pp_vir %>%
  mutate(any_rvp = as.factor(ifelse(is.na(rvp_avg_ct), "N", "Y")), .after = any_virus)
```

Create a new column that just contains a y or n for any respiratory virus categories
```{r}
data_pp_vir = data_pp_vir %>%
  mutate(rvp_any_adeno = as.factor(ifelse(is.na(rvp_adeno_ct), "N", "Y")), .before = rvp_adeno_ct) %>%
  mutate(rvp_any_rv = as.factor(ifelse(is.na(rvp_rv_ct), "N", "Y")), .before = rvp_adeno_ct) %>%
  mutate(rvp_any_hmpv = as.factor(ifelse(is.na(rvp_hmpv_ct), "N", "Y")), .before = rvp_adeno_ct) %>%
  mutate(rvp_any_cov229 = as.factor(ifelse(is.na(rvp_cov229_ct), "N", "Y")), .before = rvp_adeno_ct) %>%
  mutate(rvp_any_covnl63 = as.factor(ifelse(is.na(rvp_covnl63_ct), "N", "Y")), .before = rvp_adeno_ct) %>%
  mutate(rvp_any_flu_a = as.factor(ifelse(is.na(rvp_flu_a_ct), "N", "Y")), .before = rvp_adeno_ct) %>%
  mutate(rvp_any_piv = as.factor(ifelse(is.na(rvp_piv_ct), "N", "Y")), .before = rvp_adeno_ct) %>%
  mutate(rvp_any_rsv = as.factor(ifelse(is.na(rvp_rsv_ct), "N", "Y")), .before = rvp_adeno_ct) 
```

Create a column that contains number of rvp infections

```{r}
data_pp_vir = data_pp_vir %>% 
  ungroup() %>% mutate(rvp_n = rowSums(dplyr::select(., starts_with('rvp_any_')) == 'Y', na.rm = TRUE), .after = any_rvp) 
data_pp_vir
```

## Process bacterial

Steps:
1. Create a column that contains max/min/mean Ct -> NA Cts is NA in these
2. For each CT, generate an imputed column and a column encoding NA status:
3. Create a new column that just contains a y or n for any bacterial categories
4. create a column that contains number of bacterial infections

```{r}
names(data_filt)
```

Create a column that contains max/min/mean Ct -> NA Cts is NA in these
```{r warning=FALSE}
data_pp_bac = data_filt %>% 
  dplyr::select(study_code, any_bacteria, bac_pneu_ct, bac_influ_ct, bac_catarr_ct, albumin_ct) 

# since the columns are not numerical -> convert
data_pp_bac = data_pp_bac %>%
  # to characters
  mutate(bac_pneu_ct = as.character(bac_pneu_ct),
       bac_influ_ct = as.character(bac_influ_ct),
       bac_catarr_ct = as.character(bac_catarr_ct)) %>%
  # parse
  mutate(bac_pneu_ct = ifelse(bac_pneu_ct == "Negative", NA, as.numeric(bac_pneu_ct)),
         bac_influ_ct = ifelse(bac_influ_ct == "Negative", NA, as.numeric(bac_influ_ct)),
         bac_catarr_ct = ifelse(bac_catarr_ct == "Negative", NA, as.numeric(bac_catarr_ct)))

data_pp_bac = data_pp_bac %>%
  # average and max/min ct
  rowwise() %>%
  # calculate
  mutate(bac_avg_ct = mean(c(bac_pneu_ct, bac_influ_ct, bac_catarr_ct), na.rm = T),
         bac_max_ct = max(c(bac_pneu_ct, bac_influ_ct, bac_catarr_ct), na.rm = T),
         bac_min_ct = min(c(bac_pneu_ct, bac_influ_ct, bac_catarr_ct), na.rm = T)) %>%
  # replace - inf
  mutate(bac_avg_ct = ifelse(bac_avg_ct < 0, NA, bac_avg_ct),
       bac_max_ct = ifelse(bac_max_ct < 0, NA, bac_max_ct),
       bac_min_ct = ifelse(bac_min_ct == Inf, NA, bac_min_ct))
data_pp_bac
```

For each CT, generate an imputed column and a column encoding NA status:
```{r}
data_pp_bac = data_pp_bac %>% 
  mutate( 
    bac_pneu_ct_imp = ifelse(is.na(bac_pneu_ct), Inf, bac_pneu_ct),
    bac_pneu_ct_is_na = as.factor(ifelse(is.na(bac_pneu_ct), "Y", "N")),
    
    bac_influ_ct_imp = ifelse(is.na(bac_influ_ct), Inf, bac_influ_ct),
    bac_influ_ct_is_na = as.factor(ifelse(is.na(bac_influ_ct), "Y", "N")),
    
    bac_catarr_ct_imp = ifelse(is.na(bac_catarr_ct), Inf, bac_catarr_ct),
    bac_catarr_ct_is_na = as.factor(ifelse(is.na(bac_catarr_ct), "Y", "N"))
    ) 
```

Create a new column that just contains a y or n for any bacterial categories
```{r}
data_pp_bac = data_pp_bac %>%
  mutate(bac_any_pneu = as.factor(ifelse(is.na(bac_pneu_ct), "N", "Y")), .before = bac_pneu_ct) %>%
  mutate(bac_any_influ = as.factor(ifelse(is.na(bac_influ_ct), "N", "Y")), .before = bac_pneu_ct) %>%
  mutate(bac_any_catarr = as.factor(ifelse(is.na(bac_catarr_ct), "N", "Y")), .before = bac_pneu_ct) 
```

create a column that contains number of bacterial infections
```{r}
data_pp_bac = data_pp_bac %>% 
  ungroup() %>% mutate(bac_n = rowSums(dplyr::select(., starts_with('bac_any_')) == 'Y', na.rm = TRUE), .after = any_bacteria) 
```

# Merge all subsets together and calc virus coinfections

```{r}
data_pp_all = merge(data_pp_demo, data_pp_vir, by = "study_code")
data_pp_all = merge(data_pp_all, data_pp_sars, by = "study_code")
data_pp_all = merge(data_pp_all, data_pp_bac, by = "study_code")

data_pp_all = data_pp_all %>%
  mutate(virus_n = ifelse(any_sars == "Y", rvp_n + 1, rvp_n), .after = any_virus)
data_pp_all
```
```{r}
names(data_pp_all)
```


```{r}
write.csv(data_pp_all, file = "./data__processed/full-processed-v1.csv", row.names = F)
```



